Lynott et al. (2014) - Potential error in statistics reporting

### 2017-11-09 manuscript.Rmd
Fixed minor issue caused by earlier merge conflict

```{r echo=FALSE, fig.width=5, fig.height=7}
#  Plot H0 and H1 regions for different significance tests
source("H0H1_plot.R")
```

Figure caption: Figure 1. Illustration of null hypotheses ($H_0$) and alternative hypotheses ($H_1$) for different types of significance tests. **A)** Equivalence test: Tests the hypothesis ($H_0$) that an effect is smaller than $\Delta_{L}$ *or* larger than $\Delta_{U}$. **B)** NHST: Tests the hypothesis ($H_0$) that an effect is equal to 0. **C)** Minimal effects test: Tests the hypothesis ($H_0$) that an effect is larger than $\Delta_{L}$ *and* smaller than $\Delta_{U}$. **D)** Inferiority test: Tests the hypothesis ($H_0$) that an effect is larger than $\Delta$.

@@

Third, researchers can set the SESOI based on the sample size they are planning to collect. The amount of data you can collect limits the inferences you can make. Given the alpha level and the sample size, researchers can calculate the smallest effect size that they have sufficient power to detect^[This approach is conceptually very similar to the *power approach*, where the effect size you had 95% power to detect is calculated, and the presence of effects more extreme than this value is rejected after observing a *p*-value *larger than* 0.05 in a traditional NHST. However, Meyners (2012) explains this approach is not recommended (even though it is common) because it ignores the possibility that effects are significant *and* equivalent, and the approach does not accurately control error rates.]. This approach can be used when there are no quantitative predictions made by a theory, and no previous studies have been performed, but researchers *can* justify their sample size. In essence, this approach allows researchers to reject the presence of effects that can be reliably detected given a specific sample size. For example, a researcher who plans to perform a (two-sided) one-sample *t*-test using an alpha of 5%, based on data from 100 observations, has 90% power to detect an effect of d = 0.364. Using equivalence bounds of $\Delta_{L} = -0.364$ and $\Delta_{U} = 0.364$ in an equivalence test allows one to reject effects the study had high statistical power to detect. An equivalence test based on this approach does not answer any theoretical question (after all, the equivalence bounds are not based on any theoretical predictions) but it answers a *resource* question: Can effects be rejected that could reliably be detected given a specific sample size? In this case, concluding equivalence would suggest that effects which could reasonably be detected with a sample size of 100 can be rejected, and that future studies probably need larger sample sizes to examine a specific question. Whether or not the answer such an equivalence test provides is interesting depends on the sample sizes used in the study, and in part on how typical the chosen sample size is for the research area. An equivalence test based on 15 observations testing against a SESOI of d = 1 cannot be expected to substantially increase our knowledge, given that most effects in psychology are substantially smaller than these bounds and require larger sample sizes to be detected. By transparently reporting the effects one can detect and reject, based on the study design, researchers can communicate the information their study contributes, and provide a starting point for a discussion about what a reasonable SESOI is. 
```{r}
p <- 0.05
N <- 100
crit_d <- abs(qt(p/2, (N*2)-2))/sqrt(N/2)
```

@@

```{r include=FALSE}
source("examples_plotgrid.R")
```
```{r echo=FALSE}
#  Plot the TOST examples
plot(example.grid)
```

Figure caption: Figure 2. Effects with TOST CIs (thick lines) and NHST CIs (thin lines), NHST H0 (solid vertical line) and TOST H0 (dashed vertical lines) displayed. **A)** Example 1 - The mean difference. **B)** Example 2 - The effects included in the meta-analysis. **C)** Example 3 - The mean difference. **D)** - The proportion difference.

@@

We find that the TOST is non-significant, t(`r round(Moon$TOST_df, 2)`) = `r round(min(abs(c(Moon$TOST_t1, Moon$TOST_t2))), 2)`, p = `r round(max(Moon$TOST_p1, Moon$TOST_p2), 2)` (figure 2C). Thus, we cannot reject the possibility that the true effect really is larger than our SESOI.^[In this example, the bounds of the TOST was set based on a raw effect of 6.25%. We could have converted this raw effect into a standardized effect (Cohen's d) for the current sample, and we would have gotten the exact same result, only then in Cohen's d ($\Delta = +/- `r round(0.0625 / sqrt((sd.asian^2 + sd.control^2) / 2), 2)`)$. Since this is a replication of a previous study, we could also have chosen to calculate the Cohen's d that corresponds to the raw SESOI in the *original* study, and used this as the bounds for the replication. Now we would no longer be answering a question about a raw effect independent of the samples. Instead, we are asking whether we in the replication can reject standardized effects larger than a certain standardized effect in the original study. Since the standard deviations of the original study were only slightly higher than in the replication in this case, Cohen's d (and thus the bounds) would also be only slightly smaller ($\Delta = +/- `r round(0.0625 / sqrt((0.17^2 + 0.20^2) / 2), 2)`)$. As standard deviations of the original study compared to the replication become larger, the bounds become more narrow.] When results are neither statistically different from zero nor statistically equivalent, there is insufficient data to draw conclusions. Further studies are needed, which can be analyzed using a (small-scale) meta-analysis. The additional data will narrow the confidence interval around the observed effect, allowing us to reject the null, reject the SESOI, or both. Analogous to the large sample sizes needed to detect small effects, a lot of data is needed to reject the SESOI when very narrow bounds are used for the equivalence test (e.g., $\Delta_{L}$ = -0.1 and $\Delta_{U}$ = 0.1).

@@

As with any statistical test, the usefulness of the result of an equivalence test depends on the question we ask. The question manifests itself in the bounds we set: Is our data surprising if the true effect lies outside of the bounds? If we test against very wide bounds, observing statistical significance can hardly be considered surprising, given that most effects in psychology are small to medium (Hemphill, 2003). Examining the papers citing Lakens (2017), we see some researchers state a SESOI without providing a justification for it (Brown and colleagues, 2017; Schumann and colleagues, 2017)^[In these two cases, one could perhaps guess that benchmark values for a Cohen’s d of “medium” size was chosen. ]. An equivalence test using a SESOI of d = 0.5 might very well answer a question the researchers are interested in (for one possible justification based on minimally important differences, see Norman et al., 2003), but researchers should always provide a detailed justification of their chosen SESOI, given that this is in essence a justification for why they ask a specific question from their data.

@@

>>>>

```{r echo=FALSE, fig.width=5, fig.height=7, fig.cap=paste("put your figure caption here")}
#  Plot H0 and H1 regions for different significance tests
source("H0H1_plot.R")
```

Figure 1: Illustration of null hypotheses ($H_0$) and alternative hypotheses ($H_1$) for different types of significance tests. **A)** Equivalence test: Tests the hypothesis ($H_0$) that an effect is smaller than $\Delta_{L}$ *or* larger than $\Delta_{U}$. **B)** NHST: Tests the hypothesis ($H_0$) that an effect is equal to 0. **C)** Minimal effects test: Tests the hypothesis ($H_0$) that an effect is larger than $\Delta_{L}$ *and* smaller than $\Delta_{U}$. **D)** Inferiority test: Tests the hypothesis ($H_0$) that an effect is larger than $\Delta$.

@@

Third, researchers can set the SESOI based on the sample size they are planning to collect. The amount of data you can collect limits the inferences you can make. Given the alpha level and the sample size, researchers can calculate the smallest effect size that they have sufficient power to detect[^footnote]. This approach can be used when there are no quantitative predictions made by a theory, and no previous studies have been performed, but researchers *can* justify their sample size. In essence, this approach allows researchers to reject the presence of effects that can be reliably detected given a specific sample size. For example, a researcher who plans to perform a (two-sided) one-sample *t*-test using an alpha of 5%, based on data from 100 observations, has 90% power to detect an effect of d = 0.364. Using equivalence bounds of $\Delta_{L} = -0.364$ and $\Delta_{U} = 0.364$ in an equivalence test allows one to reject effects the study had high statistical power to detect. An equivalence test based on this approach does not answer any theoretical question (after all, the equivalence bounds are not based on any theoretical predictions) but it answers a *resource* question: Can effects be rejected that could reliably be detected given a specific sample size? In this case, concluding equivalence would suggest that effects which could reasonably be detected with a sample size of 100 can be rejected, and that future studies probably need larger sample sizes to examine a specific question. Whether or not the answer such an equivalence test provides is interesting depends on the sample sizes used in the study, and in part on how typical the chosen sample size is for the research area. An equivalence test based on 15 observations testing against a SESOI of d = 1 cannot be expected to substantially increase our knowledge, given that most effects in psychology are substantially smaller than these bounds and require larger sample sizes to be detected. By transparently reporting the effects one can detect and reject, based on the study design, researchers can communicate the information their study contributes, and provide a starting point for a discussion about what a reasonable SESOI is. 
```{r}
p <- 0.05
N <- 100
crit_d <- abs(qt(p/2, (N*2)-2))/sqrt(N/2)
```

[^footnote]: This approach is conceptually very similar to the *power approach*, where the effect size you had 95% power to detect is calculated, and the presence of effects more extreme than this value is rejected after observing a *p*-value *larger than* 0.05 in a traditional NHST. However, Meyners (2012) explains this approach is not recommended (even though it is common) because it ignores the possibility that effects are significant *and* equivalent, and the approach does not accurately control error rates.

@@

```{r include=FALSE}
source("examples_plotgrid.R")
```
```{r echo=FALSE, fig.cap=paste("Example effects plotted with TOST CIs (thick lines) and NHST CIs (thin lines), NHST H0 (solid vertical line) and TOST H0 (dashed vertical lines) displayed. **A)** Example 1 - The mean difference. **B)** Example 2 - The effects included in the meta-analysis. **C)** Example 3 - The mean difference. **D)** Example 4 - The proportion difference. **E)** Example 5 - The pearson correlation")}
#  Plot the TOST examples
plot(example.grid)
```

@@

We find that the TOST is non-significant, t(`r round(Moon$TOST_df, 2)`) = `r round(min(abs(c(Moon$TOST_t1, Moon$TOST_t2))), 2)`, p = `r round(max(Moon$TOST_p1, Moon$TOST_p2), 2)` (figure 2C). Thus, we cannot reject the possibility that the true effect really is larger than our SESOI.[^footnote] When results are neither statistically different from zero nor statistically equivalent, there is insufficient data to draw conclusions. Further studies are needed, which can be analyzed using a (small-scale) meta-analysis. The additional data will narrow the confidence interval around the observed effect, allowing us to reject the null, reject the SESOI, or both. Analogous to the large sample sizes needed to detect small effects, a lot of data is needed to reject the SESOI when very narrow bounds are used for the equivalence test (e.g., $\Delta_{L}$ = -0.1 and $\Delta_{U}$ = 0.1).

[^footnote]: In this example, the bounds of the TOST was set based on a raw effect of 6.25%. We could have converted this raw effect into a standardized effect (Cohen's d) for the current sample, and we would have gotten the exact same result, only then in Cohen's d ($\Delta = +/- `r round(0.0625 / sqrt((sd.asian^2 + sd.control^2) / 2), 2)`)$. Since this is a replication of a previous study, we could also have chosen to calculate the Cohen's d that corresponds to the raw SESOI in the *original* study, and used this as the bounds for the replication. Now we would no longer be answering a question about a raw effect independent of the samples. Instead, we are asking whether we in the replication can reject standardized effects larger than a certain standardized effect in the original study. Since the standard deviations of the original study were only slightly higher than in the replication in this case, Cohen's d (and thus the bounds) would also be only slightly smaller ($\Delta = +/- `r round(0.0625 / sqrt((0.17^2 + 0.20^2) / 2), 2)`)$. As standard deviations of the original study compared to the replication become larger, the bounds become more narrow.

@@

As with any statistical test, the usefulness of the result of an equivalence test depends on the question we ask. The question manifests itself in the bounds we set: Is our data surprising if the true effect lies outside of the bounds? If we test against very wide bounds, observing statistical significance can hardly be considered surprising, given that most effects in psychology are small to medium (Hemphill, 2003). Examining the papers citing Lakens (2017), we see some researchers state a SESOI without providing a justification for it (Brown and colleagues, 2017; Schumann and colleagues, 2017)[^footnote]. An equivalence test using a SESOI of d = 0.5 might very well answer a question the researchers are interested in (for one possible justification based on minimally important differences, see Norman et al., 2003), but researchers should always provide a detailed justification of their chosen SESOI, given that this is in essence a justification for why they ask a specific question from their data.

@@

[^footnote]: In these two cases, one could perhaps guess that benchmark values for a Cohen’s d of “medium” size was chosen.
